---
title: "Baseball Salary Analysis"
author: "Edgar Hernandez, Ryan Sims, Jessica Tomas"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    theme: readable
    toc: yes
urlcolor: cyan
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We will be examining what statistics related to a baseball player's performance have a significant effect on the salary of a player. This is particularly interesting as it is sort of the flip-side of the coin to the traditional Sabermetrics employed by the Oakland Athletics in the 1990's to analyze player statistics to try to assemble a winning team. We will be instead be leveraging the dataset to instead work for the players, hopefully determining which statistic(s) a player should focus on improving in order to increase their compensation.

Before we begin our analysis, we must first load the data using the `Lahman` library:
```{r warning=FALSE, include=TRUE}
library(Lahman)
```

We then join the Salaries, Fielding, and Batting tables:
```{r warning=FALSE, include=FALSE}
# Joining relevant tables
tables = list(Salaries, Fielding, Batting)
salary_data = Reduce(function(x,y) merge(x,y, by= c("playerID", "yearID", "teamID","lgID")), tables)
# Cleaning up redundant columns
salary_data = subset(salary_data, subset = yearID > 2000, 
                     select = -c(stint.x, stint.y, G.x, G.y, SB.x, SB.y, CS.x, CS.y, PB, WP, ZR))

# Converting character variables to factor
salary_data[sapply(salary_data, is.character)] = lapply(salary_data[sapply(salary_data, is.character)], as.factor)
```

### Methods

First, we visualize pairwise correlations between `Salary` and other variables:

- Calculating the correlations:
```{r cache = TRUE, warning = FALSE, message= FALSE}
salary_data = as.data.frame(lapply(salary_data, as.integer))
salary_data_cor = cor(as.data.frame(salary_data))
```

- Plotting the relationships:
```{r cache = TRUE, warning = FALSE, message= FALSE}
library(ggcorrplot)
my_plt = ggcorrplot(salary_data_cor, lab = TRUE, lab_size = 1, show.diag = TRUE, type = "lower")
my_plt + theme(axis.text.x = element_text(size = 5), axis.text.y = element_text(size = 5))
```

```{r}
salary_data = as.data.frame(salary_data)
baseball_fit = lm(salary ~ ., data = salary_data)
baseball_aic = step(baseball_fit, direction = "backward", trace = 0)
```

Before generating models, we being by splitting the data into a test and train dataset. For this project we will split 60% of the data into a training set and 40% of the data into a testing set.

```{r}
salary_trn_idx = sample(nrow(salary_data), size = trunc(0.60 * nrow(salary_data)))
salary_trn = salary_data[salary_trn_idx, ]
salary_tst = salary_data[-salary_trn_idx, ]
```

