---
title: "Baseball Salary Analysis - Project Proposal"
author: "Edgar Hernandez, Ryan Sims, Jessica Tomas"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
urlcolor: cyan
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description of Dataset

This dataset contains pitching, hitting, and fielding statistics for Major League Baseball from 1871
through 2021. It includes data from the two current leagues (American and National), the four other
"major" leagues (American Association, Union Association, Players League, and Federal League),
and the National Association of 1871-1875. 

```{r warning=FALSE, include=FALSE}
library(Lahman)
# Joining relevant tables
tables = list(Salaries, Fielding, Batting)
salary_data = Reduce(function(x,y) merge(x,y, by= c("playerID", "yearID", "teamID","lgID")), tables)
# Cleaning up redundant columns
salary_data = subset(salary_data, 
                     select = -c(stint.x, stint.y, G.x, G.y, SB.x, SB.y, CS.x, CS.y))
```

Details:

- The relevant portion of the dataset that we will be using contains `r length(colnames(salary_data))` variables. Some of the more critical variables are `Salary`, `Team`, `Player`, `Wins`, `Losses`, `Position`, `Games Started`, `Runs`, `Hits`, and `Homeruns`.
```{r indent = "    "}
summary(salary_data)
```



## Dataset Background Information

## Dataset Interest

## Evidence of Data

This dataset can be accessed directly in `R` through the `Lahman` library. However, we will also be providing the extracted `.csv` file for the data.
