---
title: "Baseball Salary Analysis - Project Report"
author: "Edgar Hernandez, Ryan Sims, Jessica Tomas"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
urlcolor: cyan
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

We will be examining statistics related to a baseball's player's performance and determining which statistics have a significant effect on a player's salary. Opposed to the traditional Sabermetrics employed by the Oakland Atheletics in the 1990’s to analyze player statistics to try to assemble a winning team we instead aim to leverage the dataset to work for the players. Our goal is to determine which statistic(s) a player should focus on improving in order to increase their compensation.

We will be using data through the `Lahman` library. We will join the`Fielding`, `Batting`, and `Salaries` tables in the library to generate the dataset we will use in this project. This dataset contains pitching, hitting, and fielding statistics for Major League Baseball from 1871 through 2021. It includes data from the two current leagues (American and National), the four other “major” leagues (American Association, Union Association, Players League, and Federal League), and the National Association of 1871-1875. This dataset was originally created in July 1995 by Sean Lahman and is maintained by Chris Dalzell and his team. This dataset contains data going back as far as 1871. We will however only be looking back so far as 1985 and only up to 2016, because the salary data only covers that range, and we are interested in the salaries of players. 

```{r warning=FALSE, include=FALSE}
library(Lahman)
# Joining relevant tables
tables = list(Salaries, Fielding, Batting)
salary_data = Reduce(function(x,y) merge(x,y, by= c("playerID", "yearID", "teamID","lgID")), tables)
# Cleaning up redundant columns
salary_data = subset(salary_data, 
                     select = -c(stint.x, stint.y, G.x, G.y, SB.x, SB.y, CS.x, CS.y, PB, WP, ZR))
```

The relevant portion of the dataset that we will be using contains `r length(colnames(salary_data))` variables. Critical variables in the dataset include: `Salary`, `Team`, `Player`, `Wins`, `Losses`, `Position`, `Games Started`, `Runs`, `Hits`, and `Homeruns`.

```{r}
summary(salary_data)
```

